Pangenome graph layout by Path-Guided Stochastic Gradient Descent

### 2023-09-22 manuscript/main.tex
typo
integrate Pjotr's feedback

	Here, we present a new pangenome graph layout algorithm which applies a path-guided stochastic gradient descent (PG-SGD) to move pairs of nodes in parallel with a modified HOGWILD! strategy.
	The algorithm computes the pangenome graph layout that best reflects the nucleotide sequences in the graph.
	To our knowledge, no algorithm takes into account such biological information to compute the graph layouts.
	PG-SGD can be extended in any number of dimensions.
	In the ODGI toolkit \citep{Guarracino2022}, we provide implementations for 1-dimensional (1D) and 2-dimensional (2D) layouts.

@@

	% \FIXME{@Jiajie + Niklas: Maybe address our SMT issue with your cache locality as a potential solution. Additionally just 2/3 sentences about going GPU?}
	The performance analysis shows that our 2D implementation outperforms \textit{BandageNG} when handling large, complex pangenome graphs.
	While \textit{BandageNG} was not able to deliver a layout of the whole HPRC graph within 1 week, our 2D PG-SGD calculated one within one day.
	%We expect that our implementation will be able to handle the next phase of the HPRC, 24 chromosomal pangenome graphs constructed from 300 individuals, without any problems.
	There are some possible optimization approaches for future work to further improve the performance of PG-SGD, making it possible for interactive use. 
	The data structure could be optimized to improve ache performance. Moreover, the high-degree of parallelism could be further exploited by using a GPU.  
	% As especially observed with chromosome 16 evaluation, our implementation is memory-bound.
	% The current data structure doesn't consider the memory access pattern.
	% Therefore, a CPU cache optimized implementation could make use of the massive parallelized calculations on a GPU.
	% Such efforts are currently in progress \url{https://github.com/nsmlzl/odgi/tree/layout_cache_optimized_cpu}.

@@

	S.H. acknowledges funding from the Central Innovation Programme (ZIM) for SMEs of the Federal Ministry for Economic Affairs and Energy of Germany.
	S.N. acknowledges Germany’s Excellence Strategy (CMFI), EXC-2124 and (iFIT)—EXC 2180–390900677.
	This work was supported by the BMBF-funded de.NBI Cloud within the German Network for Bioinformatics Infrastructure (de.NBI) (031A532B, 031A533A, 031A533B, 031A534A, 031A535A, 031A537A, 031A537B, 031A537C, 031A537D and 031A538A).
	A.G. acknowledges efforts by Nicole Soranzo to establish a pangenome research unit at the Human Technopole in Milan, Italy.
	JNM.S., J.L., Z.Z., P.P., and E.G. acknowledge funding from the NSF PPoSS Award \#2118709.

>>>>

	Here, we present a new pangenome graph layout algorithm which applies a path-guided stochastic gradient descent (PG-SGD) to move pairs of nodes in parallel with a modified HOGWILD! strategy.
	The algorithm computes the pangenome graph layout that best reflects the nucleotide sequences in the graph.
	To our knowledge, no algorithm takes into account such biological information to compute the graph layouts.
	PG-SGD can be extended in any number of dimensions.
	In the ODGI toolkit \citep{Guarracino2022}, we provide implementations for 1-dimensional (1D) and 2-dimensional (2D) layouts. The respective algorithms were already successfully applied to build and visualize large scale pangenome graphs of the Human Pangenome Resource Consortium (HPRC)~\citep{Liao2023, Guarracino2023}.

@@

	% \FIXME{@Jiajie + Niklas: Maybe address our SMT issue with your cache locality as a potential solution. Additionally just 2/3 sentences about going GPU?}
	The performance analysis shows that our 2D implementation outperforms \textit{BandageNG} when handling large, complex pangenome graphs.
	While \textit{BandageNG} was not able to deliver a layout of the whole HPRC graph within 1 week, our 2D PG-SGD calculated one within one day.
	%We expect that our implementation will be able to handle the next phase of the HPRC, 24 chromosomal pangenome graphs constructed from 300 individuals, without any problems.
	There are some possible optimization approaches for future work to further improve the performance of PG-SGD, making it possible for interactive use. 
	The data structure could be optimized to improve cache performance. Moreover, the high-degree of parallelism could be further exploited by using a GPU.  
	% As especially observed with chromosome 16 evaluation, our implementation is memory-bound.
	% The current data structure doesn't consider the memory access pattern.
	% Therefore, a CPU cache optimized implementation could make use of the massive parallelized calculations on a GPU.
	% Such efforts are currently in progress \url{https://github.com/nsmlzl/odgi/tree/layout_cache_optimized_cpu}.

@@

	S.H. acknowledges funding from the Central Innovation Programme (ZIM) for SMEs of the Federal Ministry for Economic Affairs and Energy of Germany.
	S.N. acknowledges Germany’s Excellence Strategy (CMFI), EXC-2124 and (iFIT)—EXC 2180–390900677.
	This work was supported by the BMBF-funded de.NBI Cloud within the German Network for Bioinformatics Infrastructure (de.NBI) (031A532B, 031A533A, 031A533B, 031A534A, 031A535A, 031A537A, 031A537B, 031A537C, 031A537D and 031A538A).
	A.G. acknowledges efforts by Nicole Soranzo to establish a pangenome research unit at the Human Technopole in Milan, Italy.
	JNM.S., J.L., Z.Z., P.P., and E.G. acknowledge funding from the NSF PPoSS Award \#2118709. The authors gratefully acknowledge support from National Institutes of Health/NIDA U01DA047638 (E.G.), National Institutes of Health/NIGMS R01GM123489 (E.G. and P.P.)
